Identity of fish fry from the “Tetí” fishery in Eastern Cuba

### 2021-09-08 analysis/paper/paper.Rmd
uploadingAppendix

```{r}
# INSTALLING PACKAGES AND LIBRARIES - if needed
packages = c("BiocManager", "raster", "ggplot2", "ggthemes", "mapproj", "here",  
             "rentrez", "ape", "seqinr", "phangorn",
             "adegenet", "pegas")
package.check <- lapply(packages, FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE, include = FALSE, warn.conflicts = FALSE, quietly = TRUE)
    }
  }
)
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

@@

The family Gobiidae includes almost 2,000 species of marine, freshwater, and brackish water fishes worldwide [@eschmeyer2020catalog]. This group is mostly represented by small and bottom-dwelling fishes found in tropical and subtropical waters. Their small size along with cryptic body colors and the paucity of clear diagnostic characters to distinguish among taxa is a substantial challenge. However, the use of molecular tools applied to species-level identification has shed light on the taxonomy of these fishes [@keith2011phylogeny; @thacker2011phylogeny; @victor2014three; @kuang2018phylogenomic] and in the last ten years, more than one hundred gobiid species have been described [@eschmeyer2020catalog].

Although most gobies inhabit marine ecosystems, some species inhabiting freshwater or estuarine ecosystems display an amphidromous lifecycle [@myers1949usage; @bell1999overview]: meaning that adults spawn in freshwater and the post-hatch larvae are carried downstream by rivers into the ocean. After spending weeks to months (depending on the species) in these marine habitats, postlarval stage individuals return to rivers. This period of postlarval return is particularly fascinating because million of them gather at the river mouth before starting their journey upstream. This migratory behavior supports artisanal fisheries worldwide [@bell1999overview; @mcdowall2007amphidromy]. Goby-fry fisheries are common practice in several countries in the Indo-Pacific [@asis2013first], the Pacific [@castellanos2011goby], and the Caribbean (@erdman1961notes; @kwaketal2016). One problem with this style of fishery is that insufficient management has led to a considerable reduction in catch [@bell1999overview]. In addition, information on the identity of the species comprising these goby-fry fisheries is still incomplete in many places [@castellanos2011goby; @asis2013first].

Gobies have been poorly studied in Cuba and available information on their identification and distribution is not conclusive [@lara2010dna]. The fish-fry “Tetí” fishery in Eastern Cuba represents a local artisanal fishery and there is no scientific information about the exploitation of this natural resource outside of the region. Based on other reports from the Caribbean (*e.g.*, @watson2000), some authors have claimed that the gobies *Sicydium plumieri* and *Sicydium punctatum* make up the “Tetí” fishery in Cuba, but evidence for those identications are lacking. “Tetí” has been historically exploited as the main ingredient of “traditional and exotic” dishes [@pichardoytapia1862; A. Hartman, pers. comm., August 01, 2011). However, recent changes to Cuban economic policies have promoted considerable growth in tourism [@rutty2019tourism], leading to increased fisheries pressure. Annual data series show increases in the number of visitors and infrastructure development in Baracoa [@baracoaStats2015; @baracoaStats2019], where this fishery takes place. Such actions along with the lack of fishing statistics might represent a threat to the conservation of these fish. This study aims to identify fish-fry species involved in the “Tetí” fishery in Eastern Cuba using molecular data.

@@

```{r tetiPhoto, fig.cap="Lateral view of goby fry “Tetí” captured in Río Yumurí, Cuba. Photo by Tomás M. Rodríguez-Cabrera."}

@@

PCR products were purified using ExoSAP-IT™ PCR Product Cleanup Reagent (Affymetrix Inc./ USB, Cleveland, OH, USA), and then sequenced on an ABI 3100 automated sequencer (Applied Biosystems). Raw sequence files were edited and aligned with the software Geneious v.9.1 [@kearse2012geneious] and submitted to GenBank (accession numbers: MZ848370– MZ848378). We used a standard nucleotide BLAST (BLASTN 2.11.0; @zhang2000greedy), optimized for highly similar sequences (MegaBLAST; @morgulis2008database) (Query IDs: 1237–1245), and recovered the identity of the Cuban sequences with 99–100% identity.

We also aimed to recover the phylogenetic position of the fish-fry samples. For this analysis, all non-Cuban sequences were retrieved from GenBank using the R package *rentrez* [@winter2017rentrez] (Appendix 1). Duplicated sequences and sequences with missing data were excluded. The dataset was edited and aligned using the R packages *seqinr* [@charif2007seqinr], *ape* [@paradis2015package], and *Biostrings* [@pages2013package]. The final alignment was obtained with the algorithm ClustalOmega [@sievers2011fast] using the R package *msa* [@bodenhofer2015msa].

@@

Freshwater gobies (Gobiiformes: Gobiidae) have been a neglected group of fishes in Cuba and most information about their diversity and ecology on the island is virtually unknown. What we do know is that given their migratory life cycle, gobies have been the target of an artisanal fishery based on postlarvae returning to rivers on the easternmost part of the island. Our results suggest that Busck’s Goby, *Sicydium buscki*, is one of the species included in the “Tetí” fishery. *Sicydium buscki* is one of the four species of amphidromous goby reported in the Greater Antilles (@watson2000; @engmanetal2019; @eschmeyer2020catalog) and it may be that these other species (*S. gilberti*, *S. plumieri*, *S. punctatum*) are part of the “Tetí” fishery but without evidence supporting an alternative we must consider *S. buscki* as the primary candidate for the identity of this fishery. Information provided by local fishermen suggests that they harvest the fry mainly from four rivers: the Duaba, Yumurí, Toa, and Miel, during the last-quarter moon phase from August to October.

@@

>>>>

```{r}
# INSTALLING PACKAGES AND LIBRARIES - if needed
packages = c("BiocManager", "raster", "ggplot2", "ggthemes", "mapproj", "here",  
             "rentrez", "ape", "seqinr", "phangorn",
             "adegenet", "pegas",
             "knitr")
package.check <- lapply(packages, FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE, include = FALSE, warn.conflicts = FALSE, quietly = TRUE)
    }
  }
)
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

@@

The family Gobiidae includes almost 2,000 species of marine, freshwater, and brackish water fishes worldwide [@eschmeyer2020catalog]. This group is mostly represented by small and bottom-dwelling fishes found in tropical and subtropical waters. Their small size along with cryptic body colors and the paucity of clear diagnostic characters to distinguish among taxa is a substantial challenge. However, the use of molecular tools applied to species-level identification has shed light on the taxonomy of these fishes [@thacker2011phylogeny; @victor2014three; @keith2017schismatogobius; @kuang2018phylogenomic] and in the last ten years, more than one hundred gobiid species have been described [@eschmeyer2020catalog].

Although most gobies inhabit marine ecosystems, some species inhabiting freshwater or estuarine ecosystems display an amphidromous lifecycle [@myers1949usage; @bell1999overview]: meaning that adults spawn in freshwater and the post-hatch larvae are carried downstream by rivers into the ocean. After spending weeks to months (depending on the species) in these marine habitats, postlarval stage individuals return to rivers. This period of postlarval return is particularly fascinating because million of them gather at the river mouth before starting their journey upstream. This migratory behavior supports artisanal fisheries worldwide [@bell1999overview; @mcdowall2007amphidromy]. Goby-fry fisheries are common practice in several countries in the Indo-Pacific [@asis2013first], the Pacific [@castellanos2011goby], and the Caribbean [@erdman1961notes; @kwaketal2016]. One problem with this style of fishery is that insufficient management has led to a considerable reduction in catch [@bell1999overview]. In addition, information on the identity of the species comprising these goby-fry fisheries is still incomplete in many places [@castellanos2011goby; @asis2013first].

Gobies have been poorly studied in Cuba and available information on their identification and distribution is not conclusive [@lara2010dna]. The fish-fry “Tetí” fishery in Eastern Cuba represents a local artisanal fishery and there is no scientific information about the exploitation of this natural resource outside of the region. Based on other reports from the Caribbean [*e.g.*, @watson2000], some authors have claimed that the gobies *Sicydium plumieri* and *Sicydium punctatum* make up the “Tetí” fishery in Cuba, but evidence for those identications are lacking. “Tetí” has been historically exploited as the main ingredient of “traditional and exotic” dishes [@pichardoytapia1862; A. Hartman, pers. comm., August 01, 2011]. However, recent changes to Cuban economic policies have promoted considerable growth in tourism [@rutty2019tourism], leading to increased fisheries pressure. Annual data series show increases in the number of visitors and infrastructure development in Baracoa [@baracoaStats2015; @baracoaStats2019], where this fishery takes place. Such actions along with the lack of fishing statistics might represent a threat to the conservation of these fish. This study aims to identify fish-fry species involved in the “Tetí” fishery in Eastern Cuba using molecular data.

@@

```{r tetiPhoto, fig.cap="Lateral view of goby fry “Tetí” captured in Río Yumurí, Cuba. Photo by Tomás M. Rodríguez-Cabrera."}
knitr::include_graphics("analysis/figures/Fig1_Rodriguez-Machado2.jpg")

@@

PCR products were purified using ExoSAP-IT™ PCR Product Cleanup Reagent (Affymetrix Inc./ USB, Cleveland, OH, USA), and then sequenced on an ABI 3100 automated sequencer (Applied Biosystems). Raw sequence files were edited and aligned with the software Geneious v.9.1 [@kearse2012geneious] and submitted to GenBank (accession numbers: MZ848370– MZ848378). We used a standard nucleotide BLAST [BLASTN 2.11.0; @zhang2000greedy], optimized for highly similar sequences [MegaBLAST; @morgulis2008database] (Query IDs: 1237–1245), and recovered the identity of the Cuban sequences with 99–100% identity.

We also aimed to recover the phylogenetic position of the fish-fry samples. For this analysis, all non-Cuban sequences were retrieved from GenBank using the R package *rentrez* [@winter2017rentrez] (Appendix 1). Duplicated sequences and sequences with missing data were excluded. The dataset was edited and aligned using the R packages *seqinr* [@charif2007seqinr], *ape* [@paradis2019ape], and *Biostrings* [@pages2013package]. The final alignment was obtained with the algorithm ClustalOmega [@sievers2011fast] using the R package *msa* [@bodenhofer2015msa].

@@

Freshwater gobies (Gobiiformes: Gobiidae) have been a neglected group of fishes in Cuba and most information about their diversity and ecology on the island is virtually unknown. What we do know is that given their migratory life cycle, gobies have been the target of an artisanal fishery based on postlarvae returning to rivers on the easternmost part of the island. Our results suggest that Busck’s Goby, *Sicydium buscki*, is one of the species included in the “Tetí” fishery. *Sicydium buscki* is one of the four species of amphidromous goby reported in the Greater Antilles [@watson2000; @engman2019puerto; @eschmeyer2020catalog] and it may be that these other species (*S. gilberti*, *S. plumieri*, *S. punctatum*) are part of the “Tetí” fishery but without evidence supporting an alternative we must consider *S. buscki* as the primary candidate for the identity of this fishery. Information provided by local fishermen suggests that they harvest the fry mainly from four rivers: the Duaba, Yumurí, Toa, and Miel, during the last-quarter moon phase from August to October.

@@

## Appendix

```{r appendix, echo = FALSE}

# This sections requires packages here and knitr [library(here), library(knitr)]

appendixCSV <- read.csv(here("/analysis/data/derived_data/Appendix_Rodriguez-Machado_et_al.csv"))

appendixDF <- data.frame(lapply(appendixCSV, function(x) {gsub("_", " ", x)}))

knitr::kable(appendixDF, caption = "Appendix 1. Species, localities, and accession number information of the COI sequences included in this study. All sequences were retrieved from GenBank. References were included when locality information is not available on GenBank. N: total number of sequences per species.")
